Host richness increases the occurrence but not the severity of bark beetle-induced tree mortality

### 2021-07-29 Code/Manuscript.Rmd
working on analyses

# Set directory structure for project
dir.create(here("Data/"), showWarnings = FALSE)
dir.create(here("Code/"), showWarnings = FALSE)
dir.create(here("Results/"), showWarnings = FALSE)
```

@@

The study area consists of subalpine lodgepole pine (*Pinus contorta*), subalpine fir (*Abies lasiocarpa*), and Engelmann spruce (*Picea engelmannii*) forest in the Intermountain West (i.e., Arizona, New Mexico, Colorado, Utah, Nevada, Idaho, Montana, Wyoming).

@@

#### Selection of plots
We selected all plots that were part of the annnual inventory and where all subplots were inventoried.

@@

```{r map study area of hosts, eval=F}
f19 <- stars::read_stars("/Volumes/GoogleDrive/My Drive/RESEARCH/GitHub/bbdiversity/Data/Spatial/ITSPM2002-subalpine/f19/w001001.adf", proxy=F)
f93 <- stars::read_stars("/Volumes/GoogleDrive/My Drive/RESEARCH/GitHub/bbdiversity/Data/Spatial/ITSPM2002-subalpine/f93/w001001.adf", proxy=F)
f108 <- stars::read_stars("/Volumes/GoogleDrive/My Drive/RESEARCH/GitHub/bbdiversity/Data/Spatial/ITSPM2002-subalpine/f108/w001001.adf", proxy=F)

@@

```{r select plots to control for sampling effort, eval=T}
#df.sub <- df[df$nplots>5, ]
df.sub <- df
## remove plots with other hosts
df.sub  <- df.sub [df.sub$BA.Otherhosts==0,]
#nrow(datBB.c)
df.sub  <- df.sub [df.sub$BA.Allhosts/df.sub$BA.total>0.75,]
```

```{r select plots with large & abundant, eval=T}
PICO.BA.threshold <- quantile(df.sub[df.sub$agentMPB=="MPB", ]$BA.PICO, 0.10)
PIEN.BA.threshold <- quantile(df.sub[df.sub$agentSB=="SB", ]$BA.PIEN, 0.10)
ABLA.BA.threshold <- quantile(df.sub[df.sub$agentWBBB=="WBBB", ]$BA.ABLA, 0.10)

PICO.QMD.threshold <- quantile(df.sub[df.sub$agentMPB=="MPB", ]$QMD.PICO, 0.10)
PIEN.QMD.threshold <- quantile(df.sub[df.sub$agentSB=="SB", ]$QMD.PIEN, 0.10)
ABLA.QMD.threshold <- quantile(df.sub[df.sub$agentWBBB=="WBBB", ]$QMD.ABLA, 0.10)
n
# PICO, PIEN & ABLA
df.sub.PICO.PIEN.ABLA <- df.sub[df.sub$BA.PICO >= PICO.BA.threshold & df.sub$BA.PIEN >= PIEN.BA.threshold & df.sub$BA.ABLA >= ABLA.BA.threshold, ]

df.sub.PICO.PIEN.ABLA <- df.sub.PICO.PIEN.ABLA[df.sub.PICO.PIEN.ABLA$QMD.PICO >= PICO.QMD.threshold & df.sub.PICO.PIEN.ABLA$QMD.PIEN >= PIEN.QMD.threshold & df.sub.PICO.PIEN.ABLA$QMD.ABLA >= ABLA.QMD.threshold, ]

df.sub.PICO.PIEN.ABLA$focalhosts2 <- "PICO, PIEN & ABLA"
df.sub.PICO.PIEN.ABLA$nhosts2 <- 3

# PICO & PIEN
df.sub.PICO.PIEN<- df.sub[df.sub$BA.PICO >= PICO.BA.threshold & df.sub$BA.PIEN >= PIEN.BA.threshold, ]

df.sub.PICO.PIEN <- df.sub.PICO.PIEN[df.sub.PICO.PIEN$QMD.PICO >= PICO.QMD.threshold & df.sub.PICO.PIEN$QMD.PIEN >= PIEN.QMD.threshold, ]

df.sub.PICO.PIEN$focalhosts2 <- "PICO & PIEN"
df.sub.PICO.PIEN < - df.sub.PICO.PIEN[!(df.sub.PICO.PIEN$PLT_CN %in% df.sub.PICO.PIEN$PLT_CN), ]
df.sub.PICO.PIEN$nhosts2 <- 2

# PICO & ABLA
df.sub.PICO.ABLA<- df.sub[df.sub$BA.PICO >= PICO.BA.threshold & df.sub$BA.ABLA >= ABLA.BA.threshold, ]

df.sub.PICO.ABLA <- df.sub.PICO.ABLA[df.sub.PICO.ABLA$QMD.PICO >= PICO.QMD.threshold & df.sub.PICO.ABLA$QMD.ABLA >= ABLA.QMD.threshold, ]

df.sub.PICO.ABLA$focalhosts2 <- "PICO & ABLA"
df.sub.PICO.ABLA$nhosts2 <- 2

df.sub.PICO.ABLA<- df.sub.PICO.ABLA[!(df.sub.PICO.ABLA$PLT_CN %in% df.sub.PICO.PIEN.ABLA$PLT_CN), ]

# PIEN & ABLA
df.sub.PIEN.ABLA<- df.sub[df.sub$BA.PIEN >= PIEN.BA.threshold & df.sub$BA.ABLA >= ABLA.BA.threshold, ]

df.sub.PIEN.ABLA <- df.sub.PIEN.ABLA[df.sub.PIEN.ABLA$QMD.PIEN >= PIEN.QMD.threshold & df.sub.PIEN.ABLA$QMD.ABLA >= ABLA.QMD.threshold, ]

df.sub.PIEN.ABLA$focalhosts2 <- "PIEN & ABLA"
df.sub.PIEN.ABLA$nhosts2 <- 2

df.sub.PIEN.ABLA <- df.sub.PIEN.ABLA[!(df.sub.PIEN.ABLA$PLT_CN %in% df.sub.PICO.PIEN.ABLA$PLT_CN), ]

###
df.focal2 <- rbind(df.sub.PICO.ABLA, df.sub.PICO.PIEN)
df.focal2 <- rbind(df.focal2, df.sub.PICO.PIEN)
df.focal2 <- rbind(df.focal2, df.sub.PICO.PIEN.ABLA)

## Affected hosts
affectedhosts2 <- df.focal2 %>% group_by(focalhosts2, affected) %>% summarise(n = n())%>% mutate(freq = prop.table(n))
affectedhosts2[affectedhosts2$affected=="affected",]
affectednhosts2 <- df.focal2 %>% group_by(nhosts2, affected) %>% summarise(n = n())%>% mutate(freq = prop.table(n))
affectednhosts2[affectednhosts2$affected=="affected",]
### MPB

MPBnhosts2 <- df.focal2 %>% group_by(nhosts2, agentMPB) %>% summarise(n = n())%>% mutate(freq = prop.table(n))

### PIEN

### ABLA

@@

```{r, child = "appendix1.Rmd"}
```

## Supplemental figures
### Comparison of stand structure and composition, climate, and regional bark beetle activity  
```{r FigStandSC,  fig.cap = "A caption", out.width = "322.56pt"}
datBB.c <-df.sub[!df.sub$focalhost=="Any",]

>>>>

# Set directory structure for project
dir.create(here("Data/"), showWarnings = FALSE)
dir.create(here("Results/"), showWarnings = FALSE)
dir.create(here("Results/Figures/"), showWarnings = FALSE)
```

@@

The study area consists of subalpine lodgepole pine (*Pinus contorta*), subalpine fir (*Abies lasiocarpa*), and Engelmann spruce (*Picea engelmannii*) forest in the Intermountain West (i.e., Arizona, New Mexico, Colorado, Utah, Nevada, Idaho, Montana, Wyoming; Fig. \@ref(fig:FigSA)

@@

#### Selection of plots
We selected all plots that were part of the annual inventory and where all subplots were inventoried.

@@

```{r map study area of hosts, eval=F}
f19 <- stars::read_stars(here("Data/Spatial/ITSPM2002-subalpine/f19/w001001.adf"), proxy=F)
f93 <- stars::read_stars(here("/Data/Spatial/ITSPM2002-subalpine/f93/w001001.adf"), proxy=F)
f108 <- stars::read_stars(here("/Data/Spatial/ITSPM2002-subalpine/f108/w001001.adf"), proxy=F)

@@

```{r select plots to control for sampling effort, eval=T}
#df.sub <- df[df$nplots>5, ]
df.sub <- df
## remove plots with other hosts
#df.sub  <- df.sub [df.sub$BA.Otherhosts==0,]
#nrow(datBB.c)
df.sub  <- df.sub [df.sub$BA.Allhosts/df.sub$BA.total>0.75,]

# select plots where focal agent is all

df.sub <- df.sub[df.sub$focalagent=="Any", ]
```

```{r select plots with large & abundant hosts, eval=T}
threshold <- 0.10
PICO.BA.threshold <- quantile(df.sub[df.sub$agentMPB=="MPB", ]$BA.PICO, threshold)
PIEN.BA.threshold <- quantile(df.sub[df.sub$agentSB=="SB", ]$BA.PIEN, threshold)
ABLA.BA.threshold <- quantile(df.sub[df.sub$agentWBBB=="WBBB", ]$BA.ABLA, threshold)

PICO.QMD.threshold <- quantile(df.sub[df.sub$agentMPB=="MPB", ]$QMD.PICO, threshold)
PIEN.QMD.threshold <- quantile(df.sub[df.sub$agentSB=="SB", ]$QMD.PIEN, threshold)
ABLA.QMD.threshold <- quantile(df.sub[df.sub$agentWBBB=="WBBB", ]$QMD.ABLA, threshold)

df.sub$hosts2.PICO <- "absent"
df.sub[df.sub$QMD.PICO >PICO.QMD.threshold & df.sub$BA.PICO >PICO.BA.threshold, "hosts2.PICO"] <- "PICO"
df.sub$hosts2.PIEN <- "absent"
df.sub[df.sub$QMD.PIEN >PIEN.QMD.threshold & df.sub$BA.PIEN >PIEN.BA.threshold, "hosts2.PIEN"] <- "PIEN"
df.sub$hosts2.ABLA<- "absent"
df.sub[df.sub$QMD.ABLA >ABLA.QMD.threshold & df.sub$BA.ABLA >ABLA.BA.threshold, "hosts2.ABLA"] <- "ABLA"

df.sub$hosts2<- NA
df.sub$hosts2<- do.call(paste, c(df.sub[c("hosts2.PICO", "hosts2.PIEN", "hosts2.ABLA")],sep="-"))
df.sub <- df.sub[!df.sub$hosts2 =='absent-absent-absent',]
df.sub$hosts2 <- factor(df.sub$hosts2, levels=c('PICO-absent-absent','absent-absent-ABLA', 'absent-PIEN-absent', 'PICO-PIEN-absent',   'PICO-absent-ABLA', 'absent-PIEN-ABLA', 'PICO-PIEN-ABLA'), labels= c('PICO', 'ABLA', 'PIEN', 'PICO & PIEN', 'PICO & ABLA', 'PIEN & ABLA', 'PICO, PIEN, & ABLA'), ordered=TRUE)
df.sub$nhosts2 <- as.numeric(factor(df.sub$hosts2, levels=c('PICO', 'ABLA', 'PIEN', 'PICO & PIEN', 'PICO & ABLA', 'PIEN & ABLA', 'PICO, PIEN, & ABLA'), labels=c(1, 1, 1, 2,2,2, 3)))

### Number of hosts

#### Any hosts
affectednhosts2 <- df.sub %>% group_by(nhosts2, affected) %>% summarise(n = n())%>% mutate(freq = prop.table(n))
affectednhosts2 <- affectednhosts2[affectednhosts2$affected=="affected",]
colnames(affectednhosts2) <- c("nhosts", "agent", "n", "freq")
affectednhosts2$agent <- "Any"

#### MPB
MPBnhosts2 <- df.sub %>% group_by(nhosts2, agentMPB) %>% summarise(n = n())%>% mutate(freq = prop.table(n))
MPBnhosts2 <- MPBnhosts2[MPBnhosts2$agentMPB=="MPB",]
colnames(MPBnhosts2) <- c("nhosts", "agent", "n", "freq")

#### SB
SBnhosts2 <- df.focal2 %>% group_by(nhosts2, agentSB) %>% summarise(n = n())%>% mutate(freq = prop.table(n))
SBnhosts2 <- SBnhosts2[SBnhosts2$agentSB=="SB",]
colnames(SBnhosts2) <- c("nhosts", "agent", "n", "freq")

#### WBBB
WBBBnhosts2 <- df.focal2 %>% group_by(nhosts2, agentWBBB) %>% summarise(n = n())%>% mutate(freq = prop.table(n))
WBBBnhosts2 <- WBBBnhosts2[WBBBnhosts2$agentWBBB=="WBBB",]
colnames(WBBBnhosts2) <- c("nhosts", "agent", "n", "freq")

#### Merge
affectednhosts2 <- rbind(as.data.frame(affectednhosts2),as.data.frame(MPBnhosts2))
affectednhosts2 <- rbind(affectednhosts2, as.data.frame(SBnhosts2))
affectednhosts2 <- rbind(affectednhosts2, as.data.frame(WBBBnhosts2))

### Host identity
#### Any hosts
affectedhosts2 <- df.sub %>% group_by(hosts2, affected) %>% summarise(n = n())%>% mutate(freq = prop.table(n))
affectedhosts2 <- affectedhosts2[affectedhosts2$affected=="affected",]
colnames(affectedhosts2) <- c("hosts", "agent", "n", "freq")
affectedhosts2$agent <- "Any"

#### MPB
MPBhosts2 <- df.sub %>% group_by(hosts2, agentMPB) %>% summarise(n = n())%>% mutate(freq = prop.table(n))
MPBhosts2 <- MPBhosts2[MPBhosts2$agentMPB=="MPB",]
colnames(MPBhosts2) <- c("hosts", "agent", "n", "freq")

#### SB
SBhosts2 <- df.sub  %>% group_by(hosts2, agentSB) %>% summarise(n = n())%>% mutate(freq = prop.table(n))
SBhosts2 <- SBhosts2[SBhosts2$agentSB=="SB",]
colnames(SBhosts2) <- c("hosts", "agent", "n", "freq")

#### WBBB
WBBBhosts2 <-df.sub  %>% group_by(hosts2, agentWBBB) %>% summarise(n = n())%>% mutate(freq = prop.table(n))
WBBBhosts2 <- WBBBhosts2[WBBBhosts2$agentWBBB=="WBBB",]
colnames(WBBBhosts2) <- c("hosts", "agent", "n", "freq")

#### Merge
affectedhosts2 <- rbind(as.data.frame(affectedhosts2),as.data.frame(MPBhosts2))
affectedhosts2 <- rbind(affectedhosts2, as.data.frame(SBhosts2))
affectedhosts2 <- rbind(affectedhosts2, as.data.frame(WBBBhosts2))
affectedhosts2$hosts <- factor(affectedhosts2$hosts, levels=c("PICO", "PIEN", "ABLA", "PICO & PIEN", "PICO & ABLA", "PIEN & ABLA", "PICO, PIEN, & ABLA"), ordered=T)

### test for significant differences
affectedhosts3 <- as.data.frame(table(df.sub$hosts2))
colnames(affectedhosts3) <- c("hosts", "N")

affectedhosts3 <- merge(affectedhosts3, affectedhosts2[affectedhosts2$agent=="Any", c("hosts", "n")], by="hosts", all=T)
colnames (affectedhosts3 ) <- c("hosts", "n", "Any")
affectedhosts3 <- merge(affectedhosts3, affectedhosts2[affectedhosts2$agent=="MPB", c("hosts", "n")], by="hosts", all=T)
colnames (affectedhosts3 ) <- c("hosts", "n", "Any", "MPB")
affectedhosts3 <- merge(affectedhosts3, affectedhosts2[affectedhosts2$agent=="SB", c("hosts", "n")], by="hosts", all=T)
colnames (affectedhosts3 ) <- c("hosts", "n", "Any", "MPB", "SB")
affectedhosts3 <- merge(affectedhosts3, affectedhosts2[affectedhosts2$agent=="WBBB", c("hosts", "n")], by="hosts", all=T)
colnames (affectedhosts3 ) <- c("hosts", "n", "Any", "MPB", "SB", "WBBB")

affectedhosts3$label <- paste0(do.call(paste, c(affectedhosts3[c("hosts", "n")],sep="\n(n=")), ")")

affectedhosts2$p.value <- NA
affectedhosts2$statistic <- NA
for(j in unique(affectedhosts2$hosts)){
  affectedhosts2j <- affectedhosts2[affectedhosts2$hosts==j, ]
  affectedhosts3j <- affectedhosts3[affectedhosts3$hosts==j,]
  for(k in unique(affectedhosts2j$agent)[!unique(affectedhosts2j$agent)=="Any"]){

    test <- prop.test(x=do.call(c,affectedhosts3j[, c("Any", k)]), n=rep(affectedhosts3j[, "n"],2), alternative="greater")
  affectedhosts2[affectedhosts2$hosts==j & affectedhosts2$agent==k, "p.value"] <- round(test$p.value,3)
    affectedhosts2[affectedhosts2$hosts==j & affectedhosts2$agent==k, "statistic"] <- test$statistic
  }
}
affectedhosts2$symbol <- ifelse( affectedhosts2$p.value<0.05, "*", "")
affectedhosts2$agent <- factor(affectedhosts2$agent, levels=c("MPB", "SB", "WBBB", "Any"), ordered=T)

affectedhosts2 <- affectedhosts2[affectedhosts2$hosts %in% c("PICO & PIEN", "PICO & ABLA", "PIEN & ABLA", "PICO, PIEN, & ABLA"), ]
affectedhosts2$hosts <- factor(affectedhosts2$hosts, levels=c("PICO & PIEN", "PICO & ABLA", "PIEN & ABLA", "PICO, PIEN, & ABLA"), labels=affectedhosts3$label[c(4,3,7,5)], ordered=T)

Fig.file <- here("Results/Figures/Figure-OccurenceGivenSuitableStands.jpg")
jpeg(Fig.file, width=fig.width2,height=fig.width1,units="in", res=300)
ggplot(affectedhosts2, aes(x=agent, y=freq, fill=agent))+geom_bar(stat="identity")+
geom_text(aes(label = symbol), vjust = 0.2)+facet_wrap(.~hosts, scales="free", nrow=1)+ylab("proportion affected")+xlab("bark beetle species")+scale_fill_manual(values=c(rev(p2col)[2:4], grey(0.25)))+theme(legend.position="none")
whatever <- dev.off()

@@

## Supplement
### Comparison of stand structure and composition, climate, and regional bark beetle activity  
```{r FigStandSC,  fig.cap = "A caption", out.width = "322.56pt"}
datBB.c <-df.sub[!df.sub$focalhost=="Any",]
